Score ReLiability? Exploring the role of score reliability for effect size heterogeneity

### 2024-05-29 Manuscript_ScoreReLiability.qmd
text updates

In light of discussions surrounding the replicability of psychological results, meta-analytic heterogeneity is understood as a crucial parameter that may inform some aspects of (non-)replicability. Heterogeneity in itself helps understand why some replications of a single phenomenon may be successful, while other are not. If heterogeneity grows, meaning the phenomenon’s effect size varies more strongly for unexplained reasons, the probability of observing an effect size around zero or even in the negative space grows larger as well. If we know the true size and heterogeneity of a phenomenon’s effect size, we would theoretically be able to establish a baseline of expected replication rate (reference).

@@

In both the initial reports of large scale replication attempts, as well as the re-analyses by Olsson-Collentine et al. (2020), standardized effect sizes (ES), such as Cohen’s d or Hedge’s g were used. For the remainder of the article, Cohen’s d, defined in equation 1, will be used as an exemplary estimate of ES, as its used across a wide range of contexts and well understood by a broad audience.

$$d = \frac{MD}{\sigma_X}$$

Here, $MD$ refers to the difference between the two groups of interest, while $\sigma_X$ is the total pooled standard deviation.

As is widely known, score reliability affects such ES. In the context of classical test (CTT) theory score reliability is defined as the ratio of true to observed score variance, as defined in equation 2.

$$\rho_{XX’} = \frac{\sigma^2_T}{\sigma^2_X}$$

In this equation $\sigma^2_T$ refers to the true variance in the sense of CTT, meaning the actual variance of the variable, undistorted by measurement error. In the same sense, $\sigma^2_X$ refers to the total variance of scores, including both the true variance and the random error variance. Lower score reliability in a data-set leads to a smaller ES, opposed to a similar data-set with higher score reliability (references). In the meta-analytic context, as score reliability is an aspect of a measuring instrument applied to a population, score reliability may not be identical across replications. If score reliability varies across replications, meaning it also carries heterogeneity, it clearly should affect heterogeneity in observed ES as well.

@@

However, if information concerning score reliability is available, it is possible to correct the individual ES for its unreliability. This process is also known as attenuation correction and, while not without its criticisms (reference), is a widespread practice in Psychology (reference). If the claims above, made by Wiernik and Dahlke or Hunter and Schmidt, were to be true, it would imply that corrections of individual observed ESs for their unreliability, any subsequently performed meta-analysis should generate lower estimates of heterogeneity. Hunter & Schmidt even claim to have derived an equation which demonstrates how heterogeneity in score reliability inflates heterogeneity in ESs. “*If the level of reliability is independent of the true effect size across studies, then, to a close approximation*:” (p. 309)

$$Var(\delta_0) = [E(a)]^2 Var(\delta) + [E(\delta)]^2 Var(a)$$

Here, $\delta_0$ refers to the “true” ES, undistorted by sampling error but not corrected for measuring error, $\delta$ refers to the “true” ES, undistorted by sampling error and measurement error and $a$ refers to score reliability in Schmidt and Hunter's notation.

@@

However, we propose that this understanding of how differences in score reliability affect the heterogeneity of observed ESs is incomplete. Hunter and Schmidt’s implication, that heterogeneity in score reliability inflates ES heterogeneity does not follow from equation 1. If equation 1 were to be true, it would imply that heterogeneity in $\delta_0$ is essentially a function where heterogeneity and mean value of $\delta$ are weighted by mean score reliability $a$ and its heterogeneity. In this scenario, it is easy to construct cases where heterogeneity in $\delta_0$ is inflated or deflated by heterogeneity in $a$. In table 1, two such cases are demonstrated. All else is held equal, only the mean score reliability $E[a]$ is changed from .7 to .8. True heterogeneity $Var[a]$ is .2, at $E[a] = .7$ this leads to a smaller $Var[\delta_0]$, while at $E[a] = .7$ this leads to a larger $Var[\delta_0]$.

@@

As Hunter & Schmidt point out, equation 1 is only valid “*If the level of reliability is independent of the true effect size across studies \[…\]”* (2014, p. 309).

However, both parameters, ES $\delta$ and score reliability $a$ in original notation can not be independent parameters by definition. ES $\delta$ is the “true” standardized mean difference, unaffected by sampling or measurement error. The ES $\delta_0$ is standardized using the total standard deviation, which is inflated by measurement error. Equation 1, demonstrating how Cohen’s d is computed demonstrates how the mean difference is standardized in this case. The ES $\delta$ on the other hand is standardized using the true standard deviation in the sense of CTT.

$$\delta = \frac{\Delta}{\sigma_T}$$

In equation 4, $\sigma_T$ is the true standard deviation, the square root of the true score variance found in equation 2. The fact that the true standard deviation parameter $\sigma_T$ is found both in equation 4 and equation 2 demonstrates that ES $\delta$ and score reliability $a$ can not be independent variables. Therefore, the basic premise of equation 3 is not fulfilled, as the assumption of variable independence could not be fulfilled.

Alternatively, assuming that both the mean differences and pooled standard deviations are no constant values across replications, meaning both carry non-zero heterogeneity, we can describe observed ESs as a random variable following a ratio distribution. Such a ratio distribution is the probability distribution, constructed by dividing one random variable by a second variable: $Z = \frac{X}{Y}$ (reference). If the distribution of the components is known, first order taylor approximation may be used to generate estimates of mean and variance of the ratio variable (reference). Using some additional assumptions, we use the Taylor-estimator of the ratio’s variance to demonstrate how differences in score reliability may affect heterogeneity in ES.

$$var[Z] \approx \frac{var[X]}{E[Y]^2} - \frac{2E[X]}{E[Y]^3} cov[X,Y] + \frac{E[X]^2}{E[Y]^4} var[Y]$$

Assuming that the random variable are uncorrelated, simplifies the equation to

$$var[Z] \approx \frac{var[X]}{E[Y]^2} + \frac{E[X]^2}{E[Y]^4} var[Y]$$

Substituting for ES, unstandardized mean difference MD and pooled standard deviation SD leads to

$$var[ES] \approx \frac{var[MD]}{E[SD]^2} + \frac{E[MD]^2}{E[SD]^4} var[SD]$$

Correcting ES for score reliability is technically done by correcting the pooled standard deviation: $SD_c = \sqrt{a V}$ . Since $a$ is always between zero and one, in the case of imperfect score reliability, $SD_c$ will always be smaller than total $SD$. Additionally, if the variance in $SD$ was introduced by differences in score reliability, these would be removed by the attenuation correction. However, there are two parameters that may introduce variance in $SD$ across replications.

Firstly, as discussed, differences in score reliability may introduce variation to SD. Secondly however, the underlying latent variable we attempted to measure may have been not identically distributed across replications. If the populations where the replications are collected from are differently spread out in terms of the variable of interest, true differences in $SD$ would remain. Re-analyses of the ManyLabs and Registered Replication Reports have shown that such differences in true variance across replications are the norm, not the exception. In that case, some variance in $SD_c$ would remain, albeit smaller than the variance in $SD$. We reformulate equation X in terms of $SD_c$

$$var[ES_c] \approx \frac{var[MD]}{E[SD_c]^2} + \frac{E[MD]^2}{E[SD_c]^4} var[SD_c]$$

The implication of this equation is, again, not straightforward. The expected value of $SD_c$ is guaranteed to be smaller than the expected value of uncorrected $SD$. Similarly, the variance in $SD_c$ is guaranteed to be smaller than the uncorrected variance in $SD$. The expected value is always placed in the denominator of the function, while the variance is placed in the numerator of the second term in equation X+1. This implies that the variance in uncorrected ES can only be larger than the variance in $ES_c$, if the reduction in mean $SD_c$ due to the attenuation correction increases the total equation result faster than the reduction in variance of $SD_c$ decreases the total equation result. To illustrate this point, we consider some parameter combinations in table 2.

@@

Table 2 demonstrates that in scenarios, where either score reliability was very high to begin with ($E[a] = .9$ in this case), leading to very little change in $E[SD]$ to $E[SD_c]$ (row 5), or where the remaining variance in $SD_c$ was comparably low (rows 3 and 4), the attenuation correction led to a reduction in ES heterogeneity. In most other scenarios, where score reliability was lower (.7 and .5 in rows 6 and 7) or remaining variance in $SD_c$ sufficiently large (row 2), the attenuation correction actually increased the heterogeneity in ES. Note that in the table, we did not manipulate expected value or variance of MD or SD.

@@

Using analytical arguments from a reformulation of equation X we have demonstrated that differences in measuring quality often do not inflate estimates of heterogeneity. Instead, assuming that MD and SD are uncorrelated, we argue that in many circumstances differences in measuring quality deflate heterogeneity in standardized mean differences. This could be understood as a form of masking, where differences in measuring quality lead to a seemingly lower heterogeneity than we would have observed without any measurement error. Clearly, these arguments are in stark contrast with claims made by Hunter & Schmidt or Wiernik & Dahlke. To further illustrate this point, we turn towards an exemplary data-set, demonstrating that we can observe the deflation in heterogeneity in empirical data.

@@

Wang et al. (2021) used several measures which may be promising for this study. Out of four measures they use to assess negative and positive emotions, three should fit conditions ii), ii) and iv); positive and negative emotional responses to ten photographs shown in the experimental trial, positive and negative state emotion after viewing the ten photographs and positive and negative anticipated emotions following the study. Over the following paragraphs, we will briefly describe the study design and measures used that are of interest to us. Details regarding the protocol, stimuli or additional measures used may be found at […].

@@

In the experimental trials, the participants viewed 10, similarly negative, photographs of the COVID-19 pandemic and rated their emotions on each photograph using Likert-style responses to the positive/negative questions. After the experimental trials, individuals responded to five negative items from the Differential Emotions Scale on fear, anger, sadness, distrust, and stress. Similarly, they responded to five positive items from the same scale, on hope, gratitude, love, inspiration, and serenity. Averages of the five positive and negative items respectively were used as measures of state emotions after the experimental trials. Similarly, negative and positive anticipated emotions were assessed, by prefacing a question of “In the next week, to what extent, if at all, do you think you will feel each of the following?” and following up with the same items used to assess state emotions. All items or the measures used may be found in the appendix XXX, or at [url…].

>>>>

In light of discussions surrounding the replicability of psychological results, meta-analytic heterogeneity is understood as a crucial parameter that may inform some aspects of (non-)replicability. Heterogeneity in itself helps understand why some replications of a single phenomenon may be successful, while other are not. If heterogeneity grows, meaning the phenomenon’s effect size varies more strongly for unexplained reasons, the probability of observing an effect size around zero or even in the negative space grows larger as well. If we know the true size and heterogeneity of a phenomenon’s effect size, we would theoretically be able to establish a baseline of expected replication rate, given no presence of publication bias or QRPs (reference).

@@

In both the initial reports of large scale replication attempts, as well as the re-analyses by Olsson-Collentine et al. (2020), standardized effect sizes (ES), such as Cohen’s d or Hedge’s g were used. For the remainder of the article, Cohen’s d, defined in equation [-@eq-SMD0], will be used as an exemplary estimate of ES, as its used across a wide range of contexts and well understood by a broad audience.

$$d = \frac{\Delta}{\sigma_X}$$ {#eq-SMD0}

Here, $\Delta$ refers to the difference in means between the two groups of interest, while $\sigma_X$ is the total pooled standard deviation.

As is widely known, score reliability affects such ES. In the context of classical test theory (CTT) score reliability is defined as the ratio of true to observed score variance, as defined in equation [-@eq-alpha].

$$a = \frac{\sigma^2_T}{\sigma^2_X}$$ {#eq-alpha}

In this equation $\sigma^2_T$ refers to the true variance in the sense of CTT, meaning the actual variance of the variable, undistorted by measurement error. In the same sense, $\sigma^2_X$ refers to the total variance of scores, including both the true variance and the random error variance. In line with Hunter & Schmidt (2014), score reliability is denoted with $a$. Lower score reliability in a data-set leads to a smaller ES, opposed to a similar data-set with identical $\Delta$ but higher score reliability (references). In the meta-analytic context, as score reliability is an aspect of a measuring instrument applied to a population, score reliability may not be identical across replications. If score reliability varies across replications, meaning it also carries heterogeneity, it clearly should affect heterogeneity in observed ES as well.

@@

However, if information concerning score reliability is available, it is possible to correct the individual ES for its unreliability. This process is also known as attenuation correction and, while not without its criticisms (reference), is a widespread practice in Psychology (reference). If the claims above, made by Wiernik and Dahlke (2020) or Hunter and Schmidt (2014), were to be true, it would imply that corrections of individual observed ESs for their unreliability, any subsequently performed meta-analysis should generate lower estimates of heterogeneity. Hunter & Schmidt even claim to have derived an equation which demonstrates how heterogeneity in score reliability inflates heterogeneity in ESs. “*If the level of reliability is independent of the true effect size across studies, then, to a close approximation*:” (2014, p. 309)

$$Var(\delta_0) = [E(a)]^2 Var(\delta) + [E(\delta)]^2 Var(a)$$ {#eq-HS}

Here, $\delta_0$ refers to the “true” ES, undistorted by sampling error but not corrected for measuring error, while $\delta$ refers to the “true” ES, undistorted by sampling error and measurement error.

@@

However, we propose that this understanding of how differences in score reliability affect the heterogeneity of observed ESs is incomplete. Hunter and Schmidt’s implication, that heterogeneity in score reliability inflates ES heterogeneity does not follow from equation [-@eq-HS]. If equation [-@eq-HS] were to be true, it would imply that heterogeneity in $\delta_0$ is essentially a function in which heterogeneity and mean value of $\delta$ are weighted by mean score reliability $a$ and its heterogeneity. In this scenario, it is easy to construct cases where heterogeneity in $\delta_0$ is inflated or deflated by heterogeneity in $a$. In table 1, two such cases are demonstrated. All else is held equal, only the mean score reliability $E[a]$ is changed from .7 to .8. True heterogeneity $Var[a]$ is .2, at $E[a] = .7$ this leads to a smaller $Var[\delta_0]$, while at $E[a] = .7$ this leads to a larger $Var[\delta_0]$.

@@

Hunter & Schmidt point out that equation [-@eq-HS] is only valid “*If the level of reliability is independent of the true effect size across studies \[…\]”* (2014, p. 309). However, both parameters, ES $\delta$ and score reliability $a$ in original notation can not, by definition, be independent parameters. ES $\delta$ is the “true” standardized mean difference, unaffected by sampling or measurement error. The ES $\delta_0$ is standardized using the total standard deviation, which is inflated by measurement error, see equation [-@eg-SMD0]. The ES $\delta$ on the other hand is standardized using the true standard deviation in the sense of CTT.

$$\delta = \frac{\Delta}{\sigma_T}$$ {#eq-SMD}

In equation [-@eq-SMD], $\sigma_T$ is the true standard deviation, the square root of the true score variance found in equation 2. The fact that the true standard deviation parameter $\sigma_T$ is found both in equation 4 and equation 2 demonstrates that ES $\delta$ and score reliability $a$ can not be independent variables. Therefore, the basic premise of equation [-@eq-HS] is not fulfilled, as the assumption of variable independence could not be fulfilled.

Alternatively, assuming that both the mean differences and pooled standard deviations are no constant values across replications, meaning both carry non-zero heterogeneity, we can describe observed ESs as a random variable following a ratio distribution. Such a ratio distribution is the probability distribution, constructed by dividing one random variable by a second variable: $Z = \frac{X}{Y}$ (reference). If the distribution of the components is known, first order taylor approximation may be used to generate estimates of mean and variance of the ratio variable (reference). Using some additional assumptions, we use the Taylor-estimator of the ratio’s variance in [-@eq-TaylorZ] to demonstrate how differences in score reliability may affect heterogeneity in ES.

$$var[Z] \approx \frac{var[X]}{E[Y]^2} - \frac{2E[X]}{E[Y]^3} cov[X,Y] + \frac{E[X]^2}{E[Y]^4} var[Y]$$ {#eq-TaylorZ}

Assuming that the random variable are uncorrelated, simplifies equation [-@eq-TaylorZ] to equation [-@eq-TaylorZs]

$$var[Z] \approx \frac{var[X]}{E[Y]^2} + \frac{E[X]^2}{E[Y]^4} var[Y]$$ {#eq-TaylorZs}

Substituting for ES, unstandardized mean difference MD and pooled standard deviation SD leads to equation [-@eq-TaylorES0]

$$var[\delta_0] \approx \frac{var[\Delta]}{E[\sigma]^2} + \frac{E[\Delta]^2}{E[\sigma]^4} var[\sigma]$$ {#eq-TaylorES0}

Correcting ES for score reliability is technically done by estimating true score standard deviation $\sigma_T$ from total score standardd eviation $\sigma$: $\sigma_T = \sqrt{a V}$.

Since $a$ is always between zero and one, in the case of imperfect score reliability, $\sigma_T$ will always be smaller than total $\sigma$. Additionally, if the variance in $\sigma$ was introduced by differences in score reliability, these would be removed by the attenuation correction. However, in total, there are two parameters that may introduce variance in $\sigma$ across replications.

Firstly, as discussed, differences in score reliability may introduce variation to $\sigma$. Secondly however, the underlying latent variable we attempted to measure may have been not identically distributed across replications. If the populations where the replications are collected from are differently spread out in terms of the variable of interest, true differences in $\sigma_T$. Re-analyses of the ManyLabs and Registered Replication Reports have shown that such differences in true variance across replications are the norm, not the exception. In that case, some variance in $\sigma_T$ would remain, albeit smaller than the variance in $\sigma$. We reformulate equation [-@eq-TaylorES0] in terms of $\sigma_T$

$$var[\delta] \approx \frac{var[\Delta]}{E[\sigma_T]^2} + \frac{E[\Delta]^2}{E[\sigma_T]^4} var[\sigma_T]$$ {#eq-TaylorES}

The implication of this equation is, again, not straightforward. The expected value of $\sigma_T$ is guaranteed to be smaller than the expected value of uncorrected $\sigma$. Similarly, the variance in $\sigma_T$ is guaranteed to be smaller than the uncorrected variance in $\sigma$. The expected value is always placed in the denominator of the function, while the variance is placed in the numerator of the second term in equation [-@eq-TaylorES]. This implies that the variance in uncorrected ES can only be larger than the variance in $\delta_c$, if the reduction in mean $\sigma_T$ due to the attenuation correction increases the total equation result faster than the reduction in variance of $\sigma_T$ decreases the total equation result. To illustrate this point, we consider some parameter combinations in table 2.

@@

Table 2 demonstrates that in scenarios, where either score reliability was very high to begin with ($E[a] = .9$ in this case), leading to very little change in $E[\sigma]$ to $E[\sigma_c]$ (row 5), or where the remaining variance in $\sigma_c$ was comparably low (rows 3 and 4), the attenuation correction led to a reduction in ES heterogeneity. In most other scenarios, where score reliability was lower (.7 and .5 in rows 6 and 7) or remaining variance in $\sigma_c$ sufficiently large (row 2), the attenuation correction actually increased the heterogeneity in ES. Note that in the table, we did not manipulate expected value or variance of $\Delta$ or $\sigma$.

@@

Using analytical arguments from a reformulation of equation X we have demonstrated that differences in measuring quality often do not inflate estimates of heterogeneity. Instead, assuming that $\Delta$ and $\sigma$ are uncorrelated, we argue that in many circumstances differences in measuring quality deflate heterogeneity in standardized mean differences. This could be understood as a form of masking, where differences in measuring quality lead to a seemingly lower heterogeneity than we would have observed without any measurement error. Clearly, these arguments are in stark contrast with claims made by Hunter & Schmidt or Wiernik & Dahlke. To further illustrate this point, we turn towards an exemplary data-set, demonstrating that we can observe the deflation in heterogeneity in empirical data.

@@

Wang et al. (2021) used several measures which may be promising for this study. Out of four measures they use to assess negative and positive emotions, three should fit conditions ii), ii) and iv); positive and negative emotional responses to ten photographs shown in the experimental trial, positive and negative state emotion after viewing the ten photographs and positive and negative anticipated emotions following the study. Over the following paragraphs, we will briefly describe the study design and measures used that are of interest to us. Details regarding the protocol, stimuli or additional measures used may be found at \[…\].

@@

In the experimental trials, the participants viewed 10, similarly negative, photographs of the COVID-19 pandemic and rated their emotions on each photograph using Likert-style responses to the positive/negative questions. After the experimental trials, individuals responded to five negative items from the Differential Emotions Scale on fear, anger, sadness, distrust, and stress. Similarly, they responded to five positive items from the same scale, on hope, gratitude, love, inspiration, and serenity. Averages of the five positive and negative items respectively were used as measures of state emotions after the experimental trials. Similarly, negative and positive anticipated emotions were assessed, by prefacing a question of “In the next week, to what extent, if at all, do you think you will feel each of the following?” and following up with the same items used to assess state emotions. All items or the measures used may be found in the appendix XXX, or at \[url…\].
